Annotated example of a spatially varying effect with sdmTMB

### 2021-11-27 doc/appendix-snow-sdmTMB.Rmd
consolidate pages some

```{r mesh, echo=TRUE, cache=FALSE}
mesh <- make_mesh(snow, xy_cols = c("X", "Y"), cutoff = 1.5)
```

```{r plot-mesh, out.width="6in"}
plot(mesh)
```

@@

```{r explore, fig.asp = 0.3}
ggplot(snow, aes(year, count, colour = nao)) + 
  geom_jitter() + 
  scale_colour_viridis_c() +
  ggsidekick::theme_sleek()
```

@@

\clearpage
We can inspect randomized quantile residuals, although this method is not guaranteed to be accurate.

@@

More accurate residuals can be generated using STAN, but this is a very slow process!
In this case, it took more than 2.5 hrs.

```{r stan-resids, eval=FALSE}
library(tmbstan)
stan_fit <- tmbstan::tmbstan(m$tmb_obj, iter = 100, chains = 1, warmup = 99)
eta <- predict(m, tmbstan_model = stan_fit)
mu <- m$family$linkinv(as.numeric(eta))
r_mcmc <- sdmTMB:::qres_nbinom2(m, y = m$data$count, mu = mu)
saveRDS(r_mcmc, "mcmc_residuals.rds")
```

@@

In order to produce pretty maps of these predictions, we back transform the spatial coordinates into the same units as our original projection and then convert back into an sf object so that the plot axis will be in latitude and longitude.

@@

```{r shapes, warning=FALSE, echo=TRUE, message=FALSE}
coast <- rnaturalearth::ne_coastline(scale = "medium", returnclass = "sf") %>% 
  sf::st_transform(crs = Albers) 
lakes <- sf::st_read(here::here("owls/ne_10m_lakes"),  quiet = TRUE)
lakes <- lakes[lakes$scalerank==0,] %>% sf::st_transform(crs = Albers) # only large lakes
```

@@

```{r p-mean, message=FALSE, echo=TRUE, eval=T}
b <- tidy(m)

p_mean <- p %>% group_by(CBCID) %>% summarise(
  X = mean(X), Y= mean(Y), 
  mean_est_count = exp(mean(est)), 
  nao_effect = exp(b[b$term == "nao", 2] + mean(zeta_s))
  )
```

```{r plot-map, echo=TRUE, eval=T, cache=FALSE}
ggplot(data = p_proj) + 
  geom_point(data = p_mean, 
     aes(X,Y, colour = nao_effect, size = mean_est_count), alpha = 0.5) +
  geom_sf(data = coast, colour = "gray50") +
  geom_sf(data = lakes, colour = "gray50", fill = NA) +
  coord_sf(xlim = c(min(p_proj$X)-50000, max(p_proj$X)-50000),# adjusts space on sides
           ylim = c(min(p_proj$Y), max(p_proj$Y))) +
  scale_colour_viridis_c(guide = guide_colourbar(
    direction = "horizontal", title.position = "top")) +
  guides(size = "none") + 
  labs(colour = "NAO effect\non Snowy Owl count") +
  theme(legend.key.height = unit(0.3, "cm"),
        legend.position = c(0.1,0.1),
        axis.title = element_blank())
```

@@

```{r plot-all-effects, echo=TRUE, eval=T, cache=FALSE}
ggplot(data = filter(p_proj, year > 1995)) + 
  geom_point(data = filter(p, year > 1995), aes(X, Y, colour = exp(est)), 
             size = 0.4, alpha = 0.5) +
  coord_sf(xlim = c(min(p_proj$X)-50000, max(p_proj$X)-50000), 
           ylim = c(min(p_proj$Y), max(p_proj$Y)))+
  scale_colour_viridis_c(trans="sqrt", # makes variation in low numbers visible 
    # can also trim the extreme high values
    limits = c(0, quantile(exp(p$est), 0.995)), na.value = "yellow",
    guide = guide_colourbar(title.position = "top")) +
  labs(colour = "Predicted counts") + 
  theme(axis.text = element_blank(), axis.ticks = element_blank()) +
  facet_wrap(~year_f)
```

Note that for all of these annually varying predictions, we are showing only the most recent 25 years.

@@

### 1. Just fixed effects

@@

```{r plot-fixed-effects, echo=TRUE, eval=T, cache=FALSE}
ggplot(data = filter(p_proj, year > 1995)) + 
  geom_point(data = filter(p, year > 1995), aes(X, Y, colour = exp(est_non_rf)), 
             size = 0.4, alpha = 0.5) +
  coord_sf(xlim = c(min(p_proj$X)-50000, max(p_proj$X)-50000), 
           ylim = c(min(p_proj$Y), max(p_proj$Y)))+
  scale_colour_viridis_c(breaks = c(0.03, 0.05, 0.07), 
    guide = guide_colourbar(title.position = "top")) +
  theme(axis.text = element_blank(), axis.ticks = element_blank()) +
  facet_wrap(~year_f)
```

### 2. Static spatial random effects

```{r plot-spatial-effects, echo=TRUE, eval=T, cache=FALSE}
ggplot(data = p_proj) + 
  geom_point(data = p, aes(X,Y, colour = omega_s), alpha = 0.5) +
  geom_sf(data = coast, colour = "gray50") +
  geom_sf(data = lakes, colour = "gray50", fill = NA) +
  coord_sf(xlim = c(min(p_proj$X)-50000, max(p_proj$X)-50000),
           ylim = c(min(p_proj$Y), max(p_proj$Y))) +
  scale_colour_viridis_c(guide = guide_colourbar(
    direction = "horizontal", title.position = "top")) +
  theme(legend.key.height = unit(0.3, "cm"),
        legend.position = c(0.1,0.1),
        axis.title = element_blank())
```

### 3. Spatiotemporal random effects

```{r plot-spatiotemporal-effects, echo=TRUE, eval=T, cache=FALSE}
ggplot(data = filter(p_proj, year > 1995)) + 
  geom_point(data = filter(p, year > 1995), aes(X, Y, colour = epsilon_st), 
             size = 0.4, alpha = 0.5) +
  coord_sf(xlim = c(min(p_proj$X)-50000, max(p_proj$X)-50000), 
           ylim = c(min(p_proj$Y), max(p_proj$Y))) +
  scale_colour_viridis_c(guide = guide_colourbar(title.position = "top")) +
  theme(axis.text = element_blank(), axis.ticks = element_blank()) +
  facet_wrap(~year_f)
```

# References

>>>>

```{r mesh, echo=TRUE, cache=FALSE, out.width="5.5in"}
mesh <- make_mesh(snow, xy_cols = c("X", "Y"), cutoff = 1.5)
plot(mesh)
```

```{r plot-mesh}

```

@@

```{r explore, fig.asp = 0.3, cache=FALSE}
ggplot(snow, aes(year, count, colour = nao)) + geom_jitter() + 
  scale_colour_viridis_c() + ggsidekick::theme_sleek()
```

@@

<!-- \clearpage -->
We can inspect randomized quantile residuals, although this method is not guaranteed to be accurate.

@@

More accurate residuals can be generated using tmbstan, but this is a very slow process!
In this case, it took more than 2.5 hrs.

```{r stan-resids, eval=FALSE}
stan_fit <- tmbstan::tmbstan(m$tmb_obj, iter = 100, chains = 1, warmup = 99)
eta <- predict(m, tmbstan_model = stan_fit)
mu <- m$family$linkinv(as.numeric(eta))
r_mcmc <- sdmTMB:::qres_nbinom2(m, y = m$data$count, mu = mu)
saveRDS(r_mcmc, "mcmc_residuals.rds")
```

@@

To produce pretty maps of these predictions, we back transformed the spatial coordinates into the same units as our original projection and then convert back into an sf object so that the plot axis will be in latitude and longitude.

@@

```{r shapes, cache=FALSE}
coast <- rnaturalearth::ne_coastline(scale = "medium", returnclass = "sf") %>% 
  sf::st_transform(crs = Albers) 
lakes <- sf::st_read(here::here("owls/ne_10m_lakes"),  quiet = TRUE)
lakes <- lakes[lakes$scalerank==0,] %>% sf::st_transform(crs = Albers) # only large lakes
```

@@

```{r p-mean, message=FALSE, echo=TRUE, eval=T, cache=FALSE}
b <- tidy(m)
p_mean <- p %>% group_by(CBCID) %>% summarise(
  X = mean(X), Y= mean(Y), 
  mean_est_count = exp(mean(est)), 
  nao_effect = exp(b[b$term == "nao", 2] + mean(zeta_s))
  )
```

```{r plot-map, echo=TRUE, eval=T, cache=FALSE}
ggplot(data = p_proj) + geom_point(data = p_mean, 
     aes(X,Y, colour = nao_effect, size = mean_est_count), alpha = 0.5) +
  geom_sf(data = coast, colour = "gray50") +
  geom_sf(data = lakes, colour = "gray50", fill = NA) +
  coord_sf(xlim = c(min(p_proj$X)-50000, max(p_proj$X)-50000),# adjusts space on sides
           ylim = c(min(p_proj$Y), max(p_proj$Y))) +
  scale_colour_viridis_c(guide = guide_colourbar(
    direction = "horizontal", title.position = "top")) +
  guides(size = "none") + labs(colour = "NAO effect\non Snowy Owl count") +
  theme(legend.position = c(0.1,0.1), axis.title = element_blank())
```

@@

```{r plot-all-effects, echo=TRUE, eval=T, cache=FALSE}
ggplot(data = filter(p_proj, year > 1978)) + 
  geom_point(data = filter(p, year > 1978), aes(X, Y, colour = exp(est)), 
             size = 0.4, alpha = 0.5) +
  coord_sf(xlim = c(min(p_proj$X)-50000, max(p_proj$X)-50000), 
           ylim = c(min(p_proj$Y), max(p_proj$Y)))+
  scale_colour_viridis_c(trans="sqrt", # makes variation in low numbers visible 
    # can also trim the extreme high values
    limits = c(0, quantile(exp(p$est), 0.995)), na.value = "yellow",
    guide = guide_colourbar(title.position = "top")) +
  labs(colour = "Predicted counts") + 
  theme(axis.text = element_blank(), axis.ticks = element_blank()) +
  facet_wrap(~year_f)
```

<!-- Note that for all of these annually varying predictions, we are showing only the most recent 25 years.  -->

@@

### 1. Static spatial random effects

```{r plot-spatial-effects, echo=TRUE, eval=T, cache=FALSE}
ggplot(data = p_proj) + 
  geom_point(data = p, aes(X,Y, colour = omega_s), alpha = 0.5) +
  geom_sf(data = coast, colour = "gray50") +
  geom_sf(data = lakes, colour = "gray50", fill = NA) +
  coord_sf(xlim = c(min(p_proj$X)-50000, max(p_proj$X)-50000),
           ylim = c(min(p_proj$Y), max(p_proj$Y))) +
  scale_colour_viridis_c(guide = guide_colourbar(
    direction = "horizontal", title.position = "top")) +
  theme(legend.position = c(0.1,0.1), axis.title = element_blank())
```

### 2. Spatiotemporal random effects

```{r plot-spatiotemporal-effects, echo=TRUE, eval=T, cache=FALSE}
ggplot(data = filter(p_proj, year > 1978)) + 
  geom_point(data = filter(p, year > 1978), aes(X, Y, colour = epsilon_st), 
             size = 0.4, alpha = 0.5) +
  coord_sf(xlim = c(min(p_proj$X)-50000, max(p_proj$X)-50000), 
           ylim = c(min(p_proj$Y), max(p_proj$Y))) +
  scale_colour_viridis_c(guide = guide_colourbar(title.position = "top")) +
  theme(axis.text = element_blank(), axis.ticks = element_blank()) +
  facet_wrap(~year_f)
```

<!-- Note that for these annually varying predictions, we are showing only the most recent 25 years.  -->

### 3. Just fixed effects

@@

```{r plot-fixed-effects, echo=TRUE, eval=T, cache=FALSE}
ggplot(data = filter(p_proj, year > 1978)) + 
  geom_point(data = filter(p, year > 1978), aes(X, Y, colour = exp(est_non_rf)), 
             size = 0.4, alpha = 0.5) +
  coord_sf(xlim = c(min(p_proj$X)-50000, max(p_proj$X)-50000), 
           ylim = c(min(p_proj$Y), max(p_proj$Y)))+
  scale_colour_viridis_c(breaks = c(0.03, 0.05, 0.07), 
    guide = guide_colourbar(title.position = "top")) +
  theme(axis.text = element_blank(), axis.ticks = element_blank()) +
  facet_wrap(~year_f)
```

\clearpage
# References
